

>---
title: "Problem Set 4"
author: "Jessica Shaw"
date: "3/26/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

# Submission

1. Fork this repository to your own account
1. Make changes to the `README.Rmd` file (including the author field above).
1. Knit the file to HTML format
1. Publish HTML on Rpubs (you will need to sign up for the free account) and *add the link below*.
1. Submit a PR with your answers.

*Link to Rpubs document:* XXX

# Overview

Take the [datacamp course on joining data](https://www.datacamp.com/courses/joining-data-in-r-with-dplyr) to learn the join verbs.

You will analyze some data in the [`nihexporter`](https://github.com/jayhesselberth/nihexporter) package to answer some questions. The questions will cover methods in `dplyr` including two-table verbs.

First you need to install the `nihexporter` package (and `devtools`, if you don't have it already).

```r
#install.packages('devtools')
#devtools::install_github("jayhesselberth/nihexporter")
```

Read the [Overview vignette](http://jayhesselberth.github.io/nihexporter/articles/nihexporter.html) to learn the structure of the package. Take a look at two tables from the package:

```{r overview}
library(nihexporter)
library(tidyverse)
library(broom)

projects
project_io
help(levels)
```
print(projects)
# A tibble: 901,059 × 13
#   application.id institute activity application.type arra.funded project.num   foa.number fiscal.year project.start
#            <int>    <fctr>   <fctr>           <fctr>      <fctr>       <chr>       <fctr>       <int>        <date>
# 1         6258248        CA      C06                1          NA C06CA091516 RFA-RR-00-02        2000    2000-09-22
# 2         6033399        RR      C06                1          NA C06RR014469   PAR-99-032        2000    2000-07-01
print(project_io)
# A tibble: 224,899 × 4
#   project.num n.pubs n.patents total.cost
#         <chr>  <int>     <int>      <dbl>
# 1  C06CA091516      1         1    1488000
# 2  C06RR013577      1         1     329812

# Problems

## Problem 1

The `projects` and `project_io` tables have the `r intersect(names(projects), names(project_io))` 
#project.num
column in common. Use this column to perform a `left_join` and identify the most expensive project for each institute. Project institutes are in the `projects` table, whereas total project costs are available in the `project_io` table. Note you will have to use the `arrange` and `slice` verbs from `dplyr`.

help(left_join)
help(arrange)
groups(projects)
projects_inst<-group_by(projects,institute)
groups(projects_inst)
attach(projects_inst)
intersect(names(projects), names(project_io))
intersect(names(projects_inst), names(project_io))
```{r problem_1, eval=FALSE}
projects %>% group_by(institute) %>% 
  left_join(., project_io, by=intersect(names(projects), names(project_io)),copy=FALSE) %>%
  select(project.num,total.cost) %>%
  arrange(desc(total.cost)) %>%
  slice(1) %>%
  arrange(desc(total.cost)) %>%
  ungroup() %>%
  slice(1) %>%
  select(project.num)
```

p1<-projects %>% group_by(institute) %>% left_join(., project_io, by=intersect(names(projects), names(project_io)),copy=FALSE) %>%
  arrange(desc(total.cost)) %>%
  slice(1) %>%
  arrange(desc(total.cost))
print(p1)

npubs<-p1 %>% select(project.num,n.pubs) %>% print()

## Problem 2

Create a plot that describes the change in output (number of publications) for the most expensive grant from the above list. Note that you cannot use `n.pubs` for this, as that is an aggregate count over time. You need a tbl of following structure, which you can then use to calculate the number of publications 
each year (hint: use `n()` in a `summarize` statement:

help(summarize)
```{r p2_tbl}
projects %>%
  head() %>%
  left_join(publinks) %>%
  left_join(publications) %>%
  select(project.num, fiscal.year, fy.cost, pmid) %>%
  na.omit()
```

  
```{r problem_2}
#begin code provided
p200<-projects %>%
#  head() %>%
#adding filter first for efficiency of joins
  filter(project.num=="ZIHLM200888") %>%
  left_join(publinks) %>%
  left_join(publications) %>%
#also adding back the variable n.pubs, stored in my npubs data frame, so I can check my results after summarizing
left_join(npubs) %>%
#code provided didn't include pub.year - must've been an oversight?
#including pub.year instead of fiscal.year
#removing fiscal.year and fy.cost because they're not necessary for the plot requested and they result in duplicates of each pmid
#duplicates of each pmid means the summarize() n() by pub.year will be incorrect
  select(project.num, pmid, pub.year, n.pubs) %>%
  na.omit() %>%
  arrange(pmid,pub.year) %>%
  unique()
#end code provided
attach(p200)

p201 <- p200 %>%
  group_by(pub.year) %>%
  summarize(n_pubs_yr=n()) %>%
  print()

# Note: Calculated number of publications (147) exceeds given value of n.pubs (126)...

#install.packages("ggthemes")
library(ggthemes)
library(ggplot2)

plot1<-ggplot(p201, aes(pub.year, n_pubs_yr)) + geom_bar(stat="identity") +
  xlab("Year") + ylab("Number of publications") + theme_bw() +
  scale_x_continuous(breaks = c(2008,2009,2010,2011,2012,2013,2014,2015),
    labels = c("2008", "2009", "2010", "2011", "2012", "2013", "2014","2015")) +
  labs(title="Publications per year of funding")  
  
plot2<-ggplot(p201, aes(x=pub.year, y=cumsum(n_pubs_yr))) + geom_line(stat="identity") +
  xlab("Year") + ylab("Number of publications") + theme_bw() +
  scale_x_continuous(breaks = c(2008,2009,2010,2011,2012,2013,2014,2015),2016,
    labels = c("2008", "2009", "2010", "2011", "2012", "2013", "2014","2015")) +
  scale_y_continuous(breaks=seq(0,150,10)) +
  labs(title="Total publications over life of grant") +
  geom_point() + geom_text(aes(label=cumsum(n_pubs_yr),hjust=0.25, vjust=1.25))

require(gridExtra)  
grid.arrange(plot1, plot2, ncol=2)

```

## Problem 3

Use the `projects` and `clinical_studies` table to calculate how many clinical studies have been started by projects from each institute.

```{r problem_3}
# First exploring the clinical_studies table
clinical_studies
arrange(clinical_studies,trial.id)
unique(clinical_studies)
# only one study status reported for each trial id and project number confirms that this is a report of studies at only one point in time
levels(clinical_studies$study.status)
## [1] "ACTIVE, NOT RECRUITING"  "AVAILABLE"               "COMPLETED"               "ENROLLING BY INVITATION" "NO LONGER AVAILABLE"    
## [6] "NOT YET RECRUITING"      "RECRUITING"              "SUSPENDED"               "TERMINATED"              "UNKNOWN STATUS"         
##[11] "WITHDRAWN"               "WITHHELD"               
# I'm not sure if any of the levels above would indicate that a project was not started...
# I'm going to assume that I should keep all statuses and just tabulate number of studies by counting trial.id.

p3<-projects %>%
  left_join(., clinical_studies, by=intersect(names(projects), names(clinical_studies)),copy=FALSE) %>%
  group_by(institute) %>%
  na.omit() %>%
  select(-application.type,-arra.funded,-study.section) %>%
  arrange(desc(project.start))
# The latest date in the project.start column is 2015, so all of the projects in this table have previously begun; 
# some are finished and some ongoing).
attach(p3)

p3 %>% 
  summarize(proj_started_by_inst=n())

```

## Problem 4

Calculate correlations between `total.cost` and `n.pubs` for each institute. You will need to `join` the `projects` and `project_io` tables to link `institute` with `total.cost` and `n.pubs`. You will then need to use `group_by` and `do` from `dplyr` and `tidy` from `broom` to calculate correlations.

```{r problem_4, eval=FALSE}

head(projects) # Contains institute, project.number
head(project_io) # Contains project.num, total.cost, n.pubs
is.numeric(project_io$n.pubs)

library(dplyr)
library(broom)

p4<-projects %>%
  dplyr::left_join(., project_io, by="project.num",copy=FALSE) %>%
  ungroup() %>%
  dplyr::group_by(institute) %>%
  select(project.num,n.pubs,total.cost) %>%
  dplyr::summarize(cor(n.pubs,total.cost,use="pairwise.complete.obs"))

# For some reason this code isn't working for me...
#p4<-projects %>%
#  dplyr::inner_join(., project_io, by="project.num",copy=FALSE) %>%
#  dplyr::group_by(institute) %>%
#  select(institute,n.pubs,total.cost) %>%
#  arrange() %>%
#  unique() %>%
#  attach() %>%
#  do(tidy(cor(n.pubs,total.cost)))
```

## Problem 5 (extra credit)

Correlate [`rcr`](https://icite.od.nih.gov/) values from the `publications` table to one of:

- total cost (overall for NIH, or by institute)
- budget mechanism (e.g., RFA, PAR, etc.)
- type of grant (e.g. R01, P01, R21, etc.). 

```{r extra_credit}
library(dplyr)
model<-projects %>%
  dplyr::left_join(., project_io, by="project.num",copy=FALSE) %>%
  ungroup() %>%
  select(-application.id, application.type,arra.funded,foa.number)

help(lme)
library(nlme)
fit<-lme(n.pubs ~ institute + total.cost, data=model, na.action=na.exclude)
summary(fit)
dplyr::summarize(cor(n.pubs,total.cost,use="pairwise.complete.obs"))
```

## Problem 6 (extra extra credit)

Use [`gganimate`](https://github.com/dgrtwo/gganimate) to illustrate the change in cost of R01s for a few institutes over time.

```{r extra_extra_credit}
```
